---
title: "Final_Report_BIOS731_homework4"
author: "Xinyu Dong"
date: "2026-04-14"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

## Problem 1: Gibbs Sampler Derivations

We observe data $y=(y_1,\dots,y_n)$, latent cluster assignments
$c=(c_1,\dots,c_n)$, and component means $\mu=(\mu_1,\dots,\mu_K)$. The
model is $$
\mu_k \sim N(0,\sigma^2), \qquad k=1,\dots,K,
$$ $$
c_i \sim \mathrm{Categorical}(1/K,\dots,1/K), \qquad i=1,\dots,n,
$$ and $$
y_i \mid c_i,\mu \sim N(\mu_{c_i},1), \qquad i=1,\dots,n.
$$

We derive a Gibbs sampler for the posterior $p(\mu,c\mid y)$.

**1. Full conditional for** $c_i$

For a fixed observation $i$, the full conditional is $$
p(c_i = k \mid y_i,\mu)
\propto p(y_i \mid c_i = k,\mu)\,p(c_i = k).
$$ Since the prior on $c_i$ is uniform, $$
p(c_i = k) = \frac{1}{K},
$$ which is constant in $k$. Therefore, $$
p(c_i = k \mid y_i,\mu)
\propto p(y_i \mid c_i = k,\mu).
$$ Under the model, $$
y_i \mid c_i = k,\mu \sim N(\mu_k,1),
$$ so $$
p(y_i \mid c_i = k,\mu)
\propto \exp\left\{-\frac{1}{2}(y_i-\mu_k)^2\right\}.
$$ Hence $$
p(c_i = k \mid y_i,\mu)
\propto \exp\left\{-\frac{1}{2}(y_i-\mu_k)^2\right\}, \qquad k=1,\dots,K.
$$ Normalizing over all $k$, $$
p(c_i = k \mid y_i,\mu)
=
\frac{
\exp\left\{-\frac{1}{2}(y_i-\mu_k)^2\right\}
}{
\sum_{\ell=1}^K
\exp\left\{-\frac{1}{2}(y_i-\mu_\ell)^2\right\}
}.
$$ Therefore, the Gibbs update is $$
c_i \mid y_i,\mu \sim \mathrm{Categorical}(p_{i1},\dots,p_{iK}),
$$ where $$
p_{ik}
=
\frac{
\exp\left\{-\frac{1}{2}(y_i-\mu_k)^2\right\}
}{
\sum_{\ell=1}^K
\exp\left\{-\frac{1}{2}(y_i-\mu_\ell)^2\right\}
}.
$$

**2. Full conditional for** $\mu_k$

Let $$
I_k = \{i : c_i = k\}, \qquad n_k = |I_k|.
$$ The prior for $\mu_k$ is $$
\mu_k \sim N(0,\sigma^2),
$$ so $$
p(\mu_k) \propto \exp\left(-\frac{\mu_k^2}{2\sigma^2}\right).
$$ The likelihood contribution from the observations assigned to cluster
$k$ is $$
\prod_{i:c_i=k} p(y_i \mid \mu_k)
\propto
\exp\left\{-\frac{1}{2}\sum_{i:c_i=k}(y_i-\mu_k)^2\right\}.
$$ Thus, $$
p(\mu_k \mid y,c)
\propto
\exp\left(-\frac{\mu_k^2}{2\sigma^2}\right)
\exp\left\{-\frac{1}{2}\sum_{i:c_i=k}(y_i-\mu_k)^2\right\}.
$$ Combining the exponents, $$
p(\mu_k \mid y,c)
\propto
\exp\left\{
-\frac{1}{2}
\left[
\frac{\mu_k^2}{\sigma^2}
+
\sum_{i:c_i=k}(y_i-\mu_k)^2
\right]
\right\}.
$$ Expand the quadratic term: $$
\sum_{i:c_i=k}(y_i-\mu_k)^2
=
\sum_{i:c_i=k}(y_i^2 - 2y_i\mu_k + \mu_k^2)
=
\sum_{i:c_i=k} y_i^2
-2\mu_k\sum_{i:c_i=k} y_i
+n_k\mu_k^2.
$$ Substituting, $$
p(\mu_k \mid y,c)
\propto
\exp\left\{
-\frac{1}{2}
\left[
\left(n_k + \frac{1}{\sigma^2}\right)\mu_k^2
-2\mu_k\sum_{i:c_i=k} y_i
+\sum_{i:c_i=k} y_i^2
\right]
\right\}.
$$ Dropping terms not involving $\mu_k$, $$
p(\mu_k \mid y,c)
\propto
\exp\left\{
-\frac{1}{2}
\left[
\left(n_k + \frac{1}{\sigma^2}\right)\mu_k^2
-2\mu_k\sum_{i:c_i=k} y_i
\right]
\right\}.
$$ This is the kernel of a normal density. Therefore, $$
\mu_k \mid y,c \sim N(m_k,V_k),
$$ where $$
V_k = \left(n_k + \frac{1}{\sigma^2}\right)^{-1},
\qquad
m_k = V_k \sum_{i:c_i=k} y_i.
$$ Equivalently, $$
\mu_k \mid y, c \sim \mathcal{N}\left(
\frac{\sum_{i:c_i=k} y_i}{\,n_k + 1/\sigma^2\,},
\;\frac{1}{\,n_k + 1/\sigma^2\,}
\right).
$$

If $n_k = 0$, then $$
\mu_k \mid y, c \sim \mathcal{N}(0, \sigma^2).
$$ **3. Gibbs sampler updates**

The Gibbs sampler alternates between the following two steps:

$$
c_i \mid y_i,\mu \sim \mathrm{Categorical}(p_{i1},\dots,p_{iK}),
$$ where

$$
p_{ik}
=
\frac{
\exp\left\{-\frac{1}{2}(y_i-\mu_k)^2\right\}
}{
\sum_{\ell=1}^K
\exp\left\{-\frac{1}{2}(y_i-\mu_\ell)^2\right\}
},
\qquad i=1,\dots,n,
$$ and

$$
\mu_k \mid y,c \sim
N\left(
\frac{\sum_{i:c_i=k} y_i}{n_k + 1/\sigma^2},
\;
\frac{1}{n_k + 1/\sigma^2}
\right),
\qquad k=1,\dots,K.
$$

Thus, at each iteration, we first update all cluster assignments
$c_1,\dots,c_n$, and then update all component means
$\mu_1,\dots,\mu_K$.

In summary, we consider a Bayesian mixture model of unit-variance
univariate Gaussians. The model is $$
\mu_k \sim N(0,\sigma^2), \qquad k=1,\dots,K,
$$ $$
c_i \sim \mathrm{Categorical}(1/K,\dots,1/K), \qquad i=1,\dots,n,
$$ and $$
y_i \mid c_i,\mu \sim N(\mu_{c_i},1), \qquad i=1,\dots,n.
$$

We derive a Gibbs sampler for the posterior distribution
$p(\mu,c \mid y)$.

For the cluster assignment $c_i$, the full conditional is $$
p(c_i = k \mid y_i,\mu)
\propto p(y_i \mid c_i = k,\mu)p(c_i = k).
$$ Since $p(c_i = k)=1/K$ is constant in $k$, this simplifies to $$
p(c_i = k \mid y_i,\mu)
\propto \exp\left\{-\frac{1}{2}(y_i-\mu_k)^2\right\}.
$$ Therefore, $$
p(c_i = k \mid y_i,\mu)
=
\frac{
\exp\left\{-\frac{1}{2}(y_i-\mu_k)^2\right\}
}{
\sum_{\ell=1}^K
\exp\left\{-\frac{1}{2}(y_i-\mu_\ell)^2\right\}
}.
$$

For the component mean $\mu_k$, let $n_k = \sum_{i=1}^n I(c_i=k)$.
Combining the normal prior and normal likelihood gives $$
\mu_k \mid y,c \sim N(m_k,V_k),
$$ where

$$
V_k = \left(n_k + \frac{1}{\sigma^2}\right)^{-1},
\qquad
m_k = V_k \sum_{i:c_i=k} y_i.
$$ Equivalently,

$$
\mu_k \mid y,c \sim
N\left(
\frac{\sum_{i:c_i=k} y_i}{n_k + 1/\sigma^2},
\;
\frac{1}{n_k + 1/\sigma^2}
\right).
$$

Thus, the Gibbs sampler alternates between:

1.  updating each $c_i$ from its categorical full conditional, and

2.  updating each $\mu_k$ from its normal full conditional.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(here)
source(here::here("source", "gibbs_k_mixture.R"))
```

The Gibbs sampler was implemented in `source/gibbs_k_mixture.R` and
called from the main R Markdown file using
`source(here::here("source", "gibbs_k_mixture.R"))`. Posterior draws for
the component means and cluster assignments were stored after burn-in,
and posterior summaries were written to `results/gibbs_k_mixture/` using
`here::here()`. Specifically, the output includes posterior means and
standard deviations for each $\mu_k$, as well as posterior cluster
probabilities and MAP cluster assignments for each observation.

```{r run-gibbs}
set.seed(1)

n <- 150
K <- 3
sigma2 <- 25

true_mu <- c(-3, 0, 4)
true_c <- sample(1:K, size = n, replace = TRUE)
y <- rnorm(n, mean = true_mu[true_c], sd = 1)

fit <- gibbs_k_mixture(
  y = y,
  K = K,
  sigma2 = sigma2,
  n_iter = 6000,
  burn_in = 1000,
  seed = 2026,
  save_output = TRUE
)

fit$summaries$mu_summary
head(fit$summaries$cluster_summary)

```
